\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-10-30 paper/paper.tex
Updated the experiments

We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}. In our experiments, this direct memory management reduced the space overhead by tens of gigabytes. When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB. Each thread tries to reuse the heap it used for the last allocation. If a heap is full or the thread cannot acquire the mutex, it moves to the next heap. With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle. Most calls to \texttt{free()} could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. \BWT{} sizes are for the native format. RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc|cc}
\hline
 & & & & & \multicolumn{2}{c}{\textbf{\BWT{} size}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} & \textbf{Unsorted} & \textbf{RLO} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB & 25.5 GB \\
     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB & 19.7 GB \\
     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB & 20.7 GB \\
     & Merged  &  771 Gbp & 7.63G &       101 bp &  129 GB & \\
\hline
\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp &      -- & 38.5 GB \\
     & TT      &  432 Gbp & 4.68G & 73 or 100 bp &      -- & 38.7 GB \\
     & AT      &  275 Gbp & 2.98G & 73 or 100 bp &      -- & 26.6 GB \\
     & TA      &  355 Gbp & 3.84G & 73 or 100 bp &      -- & 32.7 GB \\
     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp &      -- &  117 GB \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- &  225 GB \\
     & Merged  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- &  226 GB \\
     & Merged  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

\item \CEU{} contains reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892). We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files representing the same individual. Then we corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}) separately for each individual.

\item \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.

@@

We can also sort the reads by their likely positions in a reference genome. This \emph{position order} is useful, if we also want to store \emph{pairing information} for the reads. Consider a graph with reads as its nodes and edges between paired reads. If we sort the reads in position order, most edges will be close to the diagonal of the edge matrix. Such matrices are very easy to compress.

>>>>

We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}. In our experiments, this direct memory management reduced the space overhead by tens of gigabytes. When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB. Each thread tries to reuse the heap it used for the last allocation. If a heap is full or the thread cannot acquire the mutex, it moves to the next heap. With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle. Most \texttt{free()} calls could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. The amount of sequence data and the number of reads in each file, and the size of the \BWT{} in the native format and in the Read Server format. RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|cc|cc|cc}
\hline
 & & \multicolumn{2}{c|}{\textbf{Data}} & \multicolumn{2}{c|}{\textbf{Native \BWT}} & \multicolumn{2}{c}{\textbf{Read Server}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Unsorted} & \textbf{RLO} & \textbf{\BWT} & \textbf{\FMI} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &      -- \\
     & NA12891 &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &      -- \\
     & NA12892 &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &      -- \\
     & Merged  &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &      -- \\
\hline
\RS  & AA      &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB & 48.3 GB \\
     & TT      &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB & 48.4 GB \\
     & AT      &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB & 32.6 GB \\
     & TA      &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB & 40.6 GB \\
     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB &  152 GB \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &  281 GB \\
     & Merged  & 2.45 Tbp & 26.5G &      -- &         &         &         \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &  281 GB \\
     & Merged  & 2.44 Tbp & 26.5G &      -- &         &         &         \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

\item \CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892). We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files for each individual. Then we corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}) separately for each individual.

\item \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned according to the last two bases.

@@

We can also sort the reads by their likely positions in a reference genome. This \emph{position order} is useful for both compression and storing the \emph{pairing information} for the reads. Consider a graph with reads as its nodes and edges between paired reads. If we sort the reads in position order, most edges will be close to the diagonal of the edge matrix. Such matrices are very easy to compress.
